Efficient Verified Implementation of Introsort and Pdqsort

### 2020-05-04 papers/sorting/IJCAR_reviews_for_rebuttal.txt

Dear Peter Lammich,

Thank you for your submission to IJCAR 2020. The IJCAR 2020
rebuttal period will be between March 9 and March 12 (AoE).

During this time, you will have access to the current state of
your reviews and have the opportunity to submit a response.
Please keep in mind the following during this
process:

* The response must focus on any factual errors in the reviews
and any questions posed by the reviewers. It must not provide
new research results or reformulate the presentation. Try to be
as concise and to the point as possible.

* The rebuttal period is an opportunity to react to the
reviews, but not a requirement to do so. Thus, if you feel the
reviews are accurate and the reviewers have not asked any
questions, then you do not have to respond.

* The reviews are as submitted by the PC members, without any
coordination between them. Thus, there may be inconsistencies.
Furthermore, these are not the final versions of the reviews.
The reviews can later be updated to take into account the
discussions at the program committee meeting, and we may find
it necessary to solicit other outside reviews after the
rebuttal period.

* The program committee will read your responses carefully and
take this information into account during the discussions. On
the other hand, the program committee will not directly respond
to your responses, either before the program committee meeting
or in the final versions of the reviews.

* Your response will be seen by all PC members who have access
to the discussion of your paper, so please try to be polite and
constructive.

The reviews on your paper are attached to this letter. To
submit your response you should log on the EasyChair Web page
for IJCAR 2020 and select your submission on the menu.

----------------------- REVIEW 1 ---------------------
SUBMISSION: 128
TITLE: Efficient Verified Implementation of Introsort and Pdqsort
AUTHORS: Peter Lammich

----------- Overall evaluation -----------
SCORE: 2 (accept)
----- TEXT:
The paper presents an application of the Isabelle-LLVM framework: two state-of-the-art sorting algorithms (i.e., Introsort and Pdqsort) have been formally verified in the Isabelle/HOL theorem prover. The notable feature of this verification effort is that the verified algorithms can be extracted to LLVM representations, and further compiled to machine code that shows competitive performance against highly optimised implementations in the GNU C++ Standard Library. The performance is marvellous since most of the time the extracted code from a theorem prover (which is usually in a functional language) is several orders of magnitude slower than an optimised C/C++ implementations. Therefore, I believe this work is of significant importance.

The paper is well written and easy to follow. Both the proofs and the benchmark can be easily accessed online. Overall, I don't have any complaint against this paper and wish to see more follow-ups of the Isabelle-LLVM framework.

----------------------- REVIEW 2 ---------------------
SUBMISSION: 128
TITLE: Efficient Verified Implementation of Introsort and Pdqsort
AUTHORS: Peter Lammich

----------- Overall evaluation -----------
SCORE: 2 (accept)
----- TEXT:
The paper presents a verification effort for some sorting algorithms that are used in practice in C++ libraries.  The algorithms are non-trivial, and algorithms of the same kind have had many bugs in the past.  The paper can also be seen as a case study for the Isabelle Refinement Framework (by the same author).  After verification in the Isabelle Framework, the algorithms are translated to LLVM using the Isabelle-LLVM framework (also by the same author).  The generated algorithms (which are verified down to the LLVM level) behave on par with the implementations provided in libraries.

The paper is very clearly written.  It is a case study, but the author has chosen the right level of abstraction to explain the verification process and the tools used.  The results are nice, exhaustive, and convincingly presented.  My sole concern about the paper is that it is basically a case study for tools previously published.  This is however afaik the first proper case study, so I believe the paper should be accepted to IJCAR.

Minor comments:

p2.  Isn't the pivot generally left out of the recursive call?  In line 11, Algo 1, it seems it is also sorted in the second recursive call.

p2. Notation l..<h is maybe unconventional, maybe explain in a few words?

p3 : ...which leaves the pivot-element as first element of the left partition.  Could you recheck?

p3, Algo 2.  Maybe the arguments of PDQSORT on line 1 are just a cut and paste of line 3: lm and d are unused in this function.

p4. the the threshold

p8. It seems notation for assert and spec changed.  There might be a good reason, but if not, fix.

p9. The the Sepref

p14. The explanation of Fig 2. could probably made simpler to understand

p13. person hours are interesting, but lines of code would also be of interest

----------------------- REVIEW 3 ---------------------
SUBMISSION: 128
TITLE: Efficient Verified Implementation of Introsort and Pdqsort
AUTHORS: Peter Lammich

----------- Overall evaluation -----------
SCORE: 1 (weak accept)
----- TEXT:
The verification of sorting algorithms that are close to actual
implementation is fiendishly difficult. The paper manages to verify
in Isabelle/HOL the Introsort and Pdqsort algorithms, which are both
part of C++ libraries. The idea is to implement abstract functions
in Isabelle/HOL and then refine them to LLVM code. This produces
code that is correct but can also be run against benchmarks (and
it turns out the code is rather competitive). The verification is
aided by Isabelle-LLVM framework. This work is a good datapoint
for the formal verification of implemented algorithms (as opposed to
idealised algorithms shown in textbooks). I therefore recommend the
acceptance of the paper.

A slight let-down of the paper is that it assumes quite a bit of background
knowledge of the Isabelle-LLVM framework. References are given, which
is appreciated. A longer version of the paper might make the material
more "stand-alone".

------------------------------------------------------

Best wishes,

Nicolas Peltier
Viorica Sofronie-Stokkermans

### 2020-05-04 papers/sorting/main.tex
IJCAR'20 paper

  As generic sorting algorithm, the GNU C++ Standard Library implements the Introsort algorithm,
  a combination of quicksort, heapsort, and insertion sort. The Boost C++ Libraries implement Pdqsort,
  an extension of Introsort that achieves linear runtime on inputs with certain patterns.

  We verify Introsort and Pdqsort in the Isabelle LLVM verification framework,
  closely following the state-of-the-art implementations from GNU and Boost.
  On an extensive benchmark set, our verified implementations perform on par with the originals.
\end{abstract}

@@

\section{Introduction}
%   In this paper, we present verified and efficient implementations of
%   Musser's introspective sorting algorithm (Introsort)~\cite{Muss97} and its
%   extension \emph{Pattern-Defeating Quicksort} (Pdqsort)~\cite{pdqsort}.
%   Using the Isabelle Refinement Framework with its LLVM backend~\cite{La19-llvm},
%   we closely follow the implementations of Introsort and Pdqsort from
%   the GNU C++ Library (libstdc++)~\cite{libstdc++} and the Boost C++ Libraries~\cite{boost}.
%   The resulting sorting algorithms
%   are verified down to their representation in the LLVM intermediate language~\cite{LLVM-manual},
%   and perform on-par with the unverified original implementations on an extensive set of benchmarks.

  Sorting algorithms are an important part of any standard library.
  The GNU C++ Library (libstdc++)~\cite{libstdc++} implements Musser's introspective sorting algorithm (Introsort)~\cite{Muss97}.
  It is a combination of quicksort, heapsort, and insertion sort,
  which has the fast average case runtime of quicksort and the optimal $O(n\log(n))$
  worst-case runtime of heapsort.
  The Boost C++ Libraries~\cite{boost} provide a state-of-the-art implementation of
  \emph{Pattern-Defeating Quicksort} (Pdqsort)~\cite{pdqsort}, an extension of Introsort to
  achieve better performance on inputs that contain certain patterns like already sorted sequences.
  Verification of these algorithms and their state-of-the-art implementations is far from trivial,
  but turns out to be manageable when handled with adequate tools.

  Sorting algorithms in standard libraries have not always been correct.
  The Timsort~\cite{Peters02} algorithm in the Java standard library has a history of bugs\footnote{see \url{https://bugs.java.com/bugdatabase/view_bug.do?bug_id=8011944}}, the (hopefully) last of which was only found by a formal verification effort~\cite{GRBBH15}.
  Also, many real-world mergesort implementations suffered from an overflow bug~\cite{bs_flaw_blogpost}.
  Finally, LLVM's libc++~\cite{libc++} implements a different quicksort based sorting algorithm. While it may be functionally correct,
  it definitely violates the C++ standard by having a quadratic worst-case run time\footnote{See~\url{https://bugs.llvm.org/show_bug.cgi?id=20837}. This has not been fixed by January 2020.}.

  In this paper, we present efficient implementations of Introsort and Pdqsort that are verified down to their
  LLVM intermediate representation~\cite{LLVM-manual}. The verification uses the
  Isabelle Refinement Framework~\cite{LaTu12}, and its recent Isabelle-LLVM backend~\cite{La19-llvm}.
  We also report on two extensions of Isabelle-LLVM, to handle nested container data structures and to automatically
  generate C-header files to interface the generated code. Thanks to the modularity of the Isabelle Refinement Framework,
  our verified algorithms can easily be reused in larger verification projects.

@@

\section{The Introsort and Pdqsort Algorithms}
  The Introsort algorithm by Musser~\cite{Muss97} is a generic unstable sorting algorithm
  that combines the good average-case runtime of quicksort~\cite{Hoare61} with the optimal $O(n\log(n))$ worst-case complexity of heapsort~\cite{Will64}.
  The basic idea is to use quicksort as main sorting algorithm, insertion sort for small partitions, and heapsort when the recursion
  depth exceeds a given limit, usually $2\lfloor\log_2 n\rfloor$ for $n$ elements.

@@

  Algorithm~\ref{alg:introsort} shows our implementation of Introsort, which closely follows the implementation in libstdc++~\cite{libstdc++}.
  The function \textsc{introsort} sorts the slice \is$l..<h$ of the list\footnote{%
    Our formalization initially uses lists to represent the sequence of elements to be sorted, and refines them to arrays later (cf.\ Sec.\ \ref{sec:imp_impl}).
  } $xs$.
  If there is more than one element (line~\ref{l:intrs:trivial}), it initializes a depth
  counter and calls the function \textsc{introsort\_aux} (line~\ref{l:intrs:aux}), which partially sorts the list such that every element is no more than
  threshold positions away from its final position in the sorted list. The remaining sorting is then done by insertion sort (line~\ref{l:intrs:finalis}).
  The function \textsc{introsort\_aux} implements a recursive quicksort scheme: recursion stops if the slice becomes smaller than the the threshold (line~\ref{l:intrs:threshold}).
  If the maximum recursion depth is exhausted, heapsort is used to sort the slice (line~\ref{l:intrs:heapsort}).
  Otherwise, the slice is partitioned (line~\ref{l:intrs:part}), and the procedure is
  recursively invoked for the two partitions (line~\ref{l:intrs:rec1}--\ref{l:intrs:rec2}).

@@

  \begin{algorithm}
  \begin{algorithmic}[1]
    \Procedure{pdqsort}{$lm,xs,l,h,d$}
      \If{$h-l>1$}~\Call{pdqsort\_aux}{$\true,xs,l,h,\log(h-l)$}\EndIf \label{l:pdq:aux}
    \EndProcedure

@@

  Algorithm~\ref{alg:pdqsort} shows our implementation of Pdqsort.
  As for Introsort, the wrapper \textsc{pdqsort} just initializes a depth counter, and then calls the function \textsc{pdqsort\_aux}
  (line~\ref{l:pdq:aux}),
  which, in contrast to Introsort, completely sorts the list, such that no final insertion sort is necessary.
  Again, the \textsc{pdqsort\_aux} function implements a recursive quicksort scheme, however, with a few additional optimizations.
  Slices smaller than the threshold are sorted with insertion sort (line~\ref{l:pdq:threshold}). If the current slice is not the leftmost one
  of the list, as indicated by the parameter $lm$,
  the element before the start of the slice is guaranteed to be smaller than any element of the slice itself. This can be exploited to omit a comparison
  in the inner loop of insertion sort (cf.~Sec.~\ref{sec:insort}).
  If the slice is not smaller than the threshold, a pivot element is selected and moved to the front of the slice (line~\ref{l:pdq:pivotfront}).
  If the pivot is equal to the element before the current slice (line~\ref{l:pdq:eqopt}), this indicates a lot of equal elements.
  The \textsc{partition\_left} function (line~\ref{l:pdq:partleft}) will put them in the left partition, and then only the right partition needs
  to be sorted recursively (line~\ref{l:pdq:recright}).
  Otherwise, \textsc{partition\_right} (line~\ref{l:pdq:part}) places elements equal to the pivot in the right partition.
  Additionally, it returns a flag $ap$ that indicates that the slice was already partitioned.
  Next, we check for a highly unbalanced partitioning (line~\ref{l:pdq:unbal}), i.e., if one partition is less than 1/8th of the overall size.
  After encountering a certain number of highly unbalanced partitionings, Pdqsort switches to heapsort (line~\ref{l:pdq:heapsort}).
  Otherwise, it will shuffle some elements in both partitions, trying to break up patterns in the input (line~\ref{l:pdq:shuffle}).
  If the input was already partitioned wrt.\ the selected pivot (indicated by the flag $ap$),
  Pdqsort will optimistically try to sort both partitions with insertion sort (line~\ref{l:pdq:maybe}).
  However, these insertion sorts abort if they cannot sort the list with a small number of swaps, limiting the penalty for being too optimistic.
  Finally, the two partitions are recursively sorted (lines~\ref{l:pdq:rec1}--\ref{l:pdq:rec2}).

  Our implementation of Pdqsort closely follows the implementation we found in Boost~\cite{boost}.
  Again, we omitted a manual tail call optimization that LLVM does automatically.
  Moreover, for certain comparison functions, Boost's Pdqsort uses a special branch-aware partitioning algorithm~\cite{EdWe16}.
  We leave its verification to future work, but note that it will easily integrate in our existing formalization.

  While Introsort and Pdqsort are based on the same idea, this presentation focuses on the more complex Pdqsort:
  apart from the more involved \textsc{pdqsort\_aux} function,
  \textsc{pivot\_to\_front} uses Tukey's 'ninther' pivot selection~\cite{Bent93}, while Introsort uses the simpler median-of-three scheme.
  It has two partitioning algorithms used in different situations, and the \textsc{partition\_right} algorithm also
  checks for already partitioned slices. Finally, with \textsc{insort} and \textsc{maybe\_sort}, it uses two different
  versions of insertion sort.

@@

  \subsection{Specification of Sorting Algorithms}\label{sec:sortspec}
  The first step to verify a sorting algorithm is to specify the desired result.
  We specify a sorting algorithm as follows:
  \begin{lstlisting}
    sort_spec xs l h
      == assert l<=h \<and> h<=|xs|; spec xs'. xs =$_{l,h}$ xs' \<and> sorted (xs'[l..<h])
  \end{lstlisting}
%     xs =$_{l,h}$ xs'
%       == xs[0..<l]=xs'[0..<l] \<and> mset (xs[l..<h]) = mset (xs'[l..<h]) \<and> xs[h..]=xs'[h..]
  here \is$|xs|$ is the length of the list \is{xs} and
  \is{xs[I]} is the slice of the list \is{xs} for indexes in the interval \is{I}.
%   and \is{xs[l..]} is short for \is{xs[l..<|xs|${}$]}.
  The equivalence relation \is{xs =$_{l,h}$ xs'} relates lists $xs$ and $xs'$ iff they are equal outside the slice \is$l..<h$
  and $xs$ is a permutation of $xs'$.
  To simplify the presentation, we assume a linear ordering on the elements.
  Note that both C++ and our actual formalization support arbitrary weak orderings~\cite{Josu12}.

@@

  Using the above statement, and an analogous statement for Introsort, we can prove the main correctness theorem:
  \begin{theorem}\label{thm:sorting_correct}
  \begin{lstlisting}
    pdqsort xs l h <= sort_spec xs l h
              $\textbf{and}$  introsort xs l h <= sort_spec xs l h
  \end{lstlisting}
  \end{theorem}

@@

%   It contains the \emph{unguarded}
%   optimization: if parameter $G$ is false, we assume that we will hit a smaller element before underflowing the list, and thus can omit
%   the comparison $l<i$ in line~\ref{alg:insert:loophead}. In particular when element comparisons are cheap, i.e.~have cost similar to index comparisons,
%   omitting a comparison from the very inner loop can have a significant effect. Again, we show the version for Pdqsort here, where the unguarded optimization
%   is used for non-leftmost partitions. The final insertion sort of Introsort uses a similar optimization, where we switch to unguarded insertion once
%   the first threshold elements have been sorted.

  Again, we split the insertion sort algorithm into two smaller parts, which are proved separately via the following
  specification for \textsc{insert}:
  \begin{lstlisting}
    ASSERT (sorted xs[l..<i] \<and> l<=i \<and> i<|xs| \<and> (G \<or> xs[l-1] <= xs[i]));
    SPEC (%xs'. xs=$_{l,i+1}$xs' \<and> sorted xs[l..<i+1])
  \end{lstlisting}
  This captures the intuition that \textsc{insert} goes from a slice that is sorted up to index $i$ to one that is sorted up to index $i+1$.

@@

\section{Imperative Implementation}\label{sec:imp_impl}
  We have presented a refinement based approach to verify the Introsort and Pdqsort algorithms,
  including most optimizations we found in their libstdc++ and Boost implementations. However, the algorithms are still expressed
  as nondetermistic programs on functional lists and unbounded natural numbers.
  In this section, we use the Isabelle-LLVM framework~\cite{La19-llvm} to (semi-)automatically
  refine them to LLVM programs on arrays and 64 bit integers.

  \subsection{The Sepref Tool}
  The the Sepref tool~\cite{La15,La19-llvm} symbolically executes an abstract program in the \is{nres}-monad,
  keeping track of refinements for every abstract variable to a concrete representation, which may use pointers to dynamically allocated memory.
  During the symbolic execution, the tool synthesizes an imperative Isabelle-LLVM program, together with a refinement proof.
  The synthesis is automatic, but usually requires some program-specific setup and boilerplate. For a detailed discussion of Sepref and Isabelle-LLVM,
  we refer the reader to~\cite{La15,La19-llvm}.

@@

  We now describe an approach towards solving this problem for Sepref.
  Abstractly, we model an array by the type \is{\<alpha> option list}\footnote{Here, \is$\\<alpha> option = None | Some \\<alpha>$ is Isabelle's option datatype,
  and \is{the (Some x) == x} is the corresponding selector function.},
  where \is{None} means that the array does currently not own the respective element.
  The abstract indexing operation then moves the
  element from the list to the result:
  \begin{lstlisting}
    move xs i == assert (i<|xs| \and xs[i]~=None); return (the (xs[i]), xs[i:=None])
  \end{lstlisting}
%   where \is{the (Some x) = x} is the selector function of the option data type.
  As no memory is shared between the result and the array, we can show the following refinement:
  \begin{lstlisting}
    (%a i. (a[i],a), %xs i. move xs i) : oarr$^d$ \x nat$_{64}^k$ -> A \x oarr
  \end{lstlisting}
  where \is{oarr} is the relation between an array and an \is{\<alpha> option list}, and \is$A$ is the relation for the array elements.
  Note that this operation does not change the concrete array \is{a}. The movement of ownership is a purely abstract concept, which results
  in no implementation overhead.

  The transition from \is{\<alpha> list} to \is{\<alpha> option list} can typically be done in an additional refinement step,
  and thus does not obfuscate the actual correctness proofs, which are still done on plain \is{\<alpha> list}.
  Moreover, the \is{\<alpha> option list} representation is only required for subroutines where extracted array elements are actually visible.
  For example, we define an operation to compare two array elements:
  \begin{lstlisting}
    cmp_idxs xs i j == assert (i<|xs| \<and> j<|xs|); return xs[i] < xs[j]
  \end{lstlisting}
  Inside this operation, we have to temporarily extract the elements $i$ and $j$ from the array,
  requiring an intermediate refinement step to \is{\<alpha> option list}.
  However, at the start and end of this operation, the array owns all its elements. For the whole operation, we thus get
  a refinement on plain arrays:
  \begin{lstlisting}
    (cmp_impl, cmp_idxs) : arr$^k$ \x nat$_{64}^k$ \x nat$_{64}^k$ -> bool
  \end{lstlisting}
  In our case, we only have to explicitly refine \textsc{insert} and \textsc{sift\_down}.
  The other subroutines use \is{cmp_idxs} and \is{swap} operations on plain lists.

@@

\section{Benchmarks}
%   We have verified the libstdc++ implementation of Introsort, and the Boost implementation of Pdqsort.
%   We benchmarked our verified implementations against the unverified originals.
%   As we did not verify the C++ code directly, re-implemented the algorithm using Isabelle-LLVM, we have to
%   test how close we came to the performance of the original.

@@

  We sorted integer arrays with $n=10^8$ elements, and string arrays with $n=10^7$ elements.
  For strings, all implementations use the same data structure and compare function.
  For integers, we disable Pdqsorts branch-aware partitioning, which we have not yet verified.
  For strings, it does not apply anyway.

>>>>

  As generic sorting algorithm, the GNU C++ Standard Library implements the introsort algorithm,
  a combination of quicksort, heapsort, and insertion sort. The Boost C++ Libraries implement pdqsort,
  an extension of introsort that achieves linear runtime on inputs with certain patterns.

  We verify introsort and pdqsort in the Isabelle LLVM verification framework,
  closely following the state-of-the-art implementations from GNU and Boost.
  On an extensive benchmark set, our verified implementations perform on par with the originals.
\end{abstract}

@@

\section{Introduction}
%   In this paper, we present verified and efficient implementations of
%   Musser's introspective sorting algorithm (introsort)~\cite{Muss97} and its
%   extension \emph{pattern-defeating quicksort} (pdqsort)~\cite{pdqsort}.
%   Using the Isabelle Refinement Framework with its LLVM backend~\cite{La19-llvm},
%   we closely follow the implementations of introsort and pdqsort from
%   the GNU C++ Library (libstdc++)~\cite{libstdc++} and the Boost C++ Libraries~\cite{boost}.
%   The resulting sorting algorithms
%   are verified down to their representation in the LLVM intermediate language~\cite{LLVM-manual},
%   and perform on-par with the unverified original implementations on an extensive set of benchmarks.

  Sorting algorithms are an important part of any standard library.
  The GNU C++ Library (libstdc++)~\cite{libstdc++} implements Musser's introspective sorting algorithm (introsort)~\cite{Muss97}.
  It is a combination of quicksort, heapsort, and insertion sort,
  which has the fast average case runtime of quicksort and the optimal $O(n\log(n))$
  worst-case runtime of heapsort.
  The Boost C++ Libraries~\cite{boost} provide a state-of-the-art implementation of
  \emph{pattern-defeating quicksort} (pdqsort)~\cite{pdqsort}, an extension of introsort to
  achieve better performance on inputs that contain certain patterns like already sorted sequences.
  Verification of these algorithms and their state-of-the-art implementations is far from trivial,
  but turns out to be manageable when handled with adequate tools.

  Sorting algorithms in standard libraries have not always been correct.
  The timsort~\cite{Peters02} algorithm in the Java standard library has a history of bugs\footnote{see \url{https://bugs.java.com/bugdatabase/view_bug.do?bug_id=8011944}}, the (hopefully) last of which was only found by a formal verification effort~\cite{GRBBH15}.
  Also, many real-world mergesort implementations suffered from an overflow bug~\cite{bs_flaw_blogpost}.
  Finally, LLVM's libc++~\cite{libc++} implements a different quicksort based sorting algorithm. While it may be functionally correct,
  it definitely violates the C++ standard by having a quadratic worst-case run time\footnote{See~\url{https://bugs.llvm.org/show_bug.cgi?id=20837}. This has not been fixed by April 2020.}.

  In this paper, we present efficient implementations of introsort and pdqsort that are verified down to their
  LLVM intermediate representation~\cite{LLVM-manual}. The verification uses the
  Isabelle Refinement Framework~\cite{LaTu12}, and its recent Isabelle-LLVM backend~\cite{La19-llvm}.
  We also report on two extensions of Isabelle-LLVM, to handle nested container data structures and to automatically
  generate C-header files to interface the generated code. Thanks to the modularity of the Isabelle Refinement Framework,
  our verified algorithms can easily be reused in larger verification projects.

@@

\section{The Introsort and Pdqsort Algorithms}
  The introsort algorithm by Musser~\cite{Muss97} is a generic unstable sorting algorithm
  that combines the good average-case runtime of quicksort~\cite{Hoare61} with the optimal $O(n\log(n))$ worst-case complexity of heapsort~\cite{Will64}.
  The basic idea is to use quicksort as main sorting algorithm, insertion sort for small partitions, and heapsort when the recursion
  depth exceeds a given limit, usually $2\lfloor\log_2 n\rfloor$ for $n$ elements.

@@

  Algorithm~\ref{alg:introsort} shows our implementation of introsort, which closely follows the implementation in libstdc++~\cite{libstdc++}.
  The function \textsc{introsort} sorts the slice from index \is$l$ (inclusive) up to index \is{h} (exclusive) of the list\footnote{%
    Our formalization initially uses lists to represent the sequence of elements to be sorted, and refines them to arrays later (cf.\ Sec.\ \ref{sec:imp_impl}).
  } $xs$.
  If there is more than one element (line~\ref{l:intrs:trivial}), it initializes a depth
  counter and calls the function \textsc{introsort\_aux} (line~\ref{l:intrs:aux}), which partially sorts the list such that every element is no more than
  threshold positions away from its final position in the sorted list. The remaining sorting is then done by insertion sort (line~\ref{l:intrs:finalis}).
  The function \textsc{introsort\_aux} implements a recursive quicksort scheme: recursion stops if the slice becomes smaller than the threshold (line~\ref{l:intrs:threshold}).
  If the maximum recursion depth is exhausted, heapsort is used to sort the slice (line~\ref{l:intrs:heapsort}).
  Otherwise, the slice is partitioned (line~\ref{l:intrs:part}), and the procedure is
  recursively invoked for the two partitions (line~\ref{l:intrs:rec1}--\ref{l:intrs:rec2}).
  Here, \textsc{partition\_pivot} moves the pivot element to the first element of the left partition, and returns the start index of the right partition.

@@

  \begin{algorithm}
  \begin{algorithmic}[1]
    \Procedure{pdqsort}{$xs,l,h$}
      \If{$h-l>1$}~\Call{pdqsort\_aux}{$\true,xs,l,h,\log(h-l)$}\EndIf \label{l:pdq:aux}
    \EndProcedure

@@

  Algorithm~\ref{alg:pdqsort} shows our implementation of pdqsort.
  As for introsort, the wrapper \textsc{pdqsort} just initializes a depth counter, and then calls the function \textsc{pdqsort\_aux}
  (line~\ref{l:pdq:aux}),
  which, in contrast to introsort, completely sorts the list, such that no final insertion sort is necessary.
  Again, the \textsc{pdqsort\_aux} function implements a recursive quicksort scheme, however, with a few additional optimizations.
  Slices smaller than the threshold are sorted with insertion sort (line~\ref{l:pdq:threshold}). If the current slice is not the leftmost one
  of the list, as indicated by the parameter $lm$,
  the element before the start of the slice is guaranteed to be smaller than any element of the slice itself. This can be exploited to omit a comparison
  in the inner loop of insertion sort (cf.~Sec.~\ref{sec:insort}).
  If the slice is not smaller than the threshold, a pivot element is selected and moved to the front of the slice (line~\ref{l:pdq:pivotfront}).
  If the pivot is equal to the element before the current slice (line~\ref{l:pdq:eqopt}), this indicates a lot of equal elements.
  The \textsc{partition\_left} function (line~\ref{l:pdq:partleft}) will put them in the left partition, and then only the right partition needs
  to be sorted recursively (line~\ref{l:pdq:recright}).
  Otherwise, \textsc{partition\_right} (line~\ref{l:pdq:part}) places elements equal to the pivot in the right partition.
  Additionally, it returns a flag $ap$ that indicates that the slice was already partitioned.
  Next, we check for a highly unbalanced partitioning (line~\ref{l:pdq:unbal}), i.e., if one partition is less than 1/8th of the overall size.
  After encountering a certain number of highly unbalanced partitionings, pdqsort switches to heapsort (line~\ref{l:pdq:heapsort}).
  Otherwise, it will shuffle some elements in both partitions, trying to break up patterns in the input (line~\ref{l:pdq:shuffle}).
  If the input was already partitioned wrt.\ the selected pivot (indicated by the flag $ap$),
  pdqsort will optimistically try to sort both partitions with insertion sort (line~\ref{l:pdq:maybe}).
  However, these insertion sorts abort if they cannot sort the list with a small number of swaps, limiting the penalty for being too optimistic.
  Finally, the two partitions are recursively sorted (lines~\ref{l:pdq:rec1}--\ref{l:pdq:rec2}).

  Our implementation of pdqsort closely follows the implementation we found in Boost~\cite{boost}.
  Again, we omitted a manual tail call optimization that LLVM does automatically.
  Moreover, for certain comparison functions, Boost's pdqsort uses a special branch-aware partitioning algorithm~\cite{EdWe16}.
  We leave its verification to future work, but note that it will easily integrate in our existing formalization.

  While introsort and pdqsort are based on the same idea, this presentation focuses on the more complex pdqsort:
  apart from the more involved \textsc{pdqsort\_aux} function,
  \textsc{pivot\_to\_front} uses Tukey's 'ninther' pivot selection~\cite{Bent93}, while introsort uses the simpler median-of-three scheme.
  It has two partitioning algorithms used in different situations, and the \textsc{partition\_right} algorithm also
  checks for already partitioned slices. Finally, with \textsc{insort} and \textsc{maybe\_sort}, it uses two different
  versions of insertion sort.

@@

  \subsection{Specification of Sorting Algorithms}\label{sec:sortspec}
  The first step to verify a sorting algorithm is to specify the desired result.
  We specify a sorting algorithm as follows:
  \begin{lstlisting}
    sort_spec xs l h
      == assert l<=h \<and> h<=|xs|; spec xs'. xs =$_{l,h}$ xs' \<and> sorted (xs'[l..<h])
  \end{lstlisting}
%     xs =$_{l,h}$ xs'
%       == xs[0..<l]=xs'[0..<l] \<and> mset (xs[l..<h]) = mset (xs'[l..<h]) \<and> xs[h..]=xs'[h..]
  here \is$|xs|$ is the length of the list \is{xs} and
  \is{xs[I]} is the slice of the list \is{xs} for indexes in the interval \is{I}\footnote{%
    An interval from index \is{l} (inclusive) to \is{h} (exclusive) is denoted as \is{l..<h}.
    If both indexes are exclusive, we write \is{l<..<h}.
  }.
%   and \is{xs[l..]} is short for \is{xs[l..<|xs|${}$]}.
  The equivalence relation \is{xs =$_{l,h}$ xs'} relates lists $xs$ and $xs'$ iff they are equal outside the slice \is$l..<h$
  and $xs$ is a permutation of $xs'$.
  To simplify the presentation, we assume a linear ordering on the elements.
  Note that both C++ and our actual formalization support arbitrary weak orderings~\cite{Josu12}.

@@

  Using the above statement, and an analogous statement for introsort, we can prove the main correctness theorem:
  \begin{theorem}\label{thm:sorting_correct}
  \begin{lstlisting}
    pdqsort xs l h <= sort_spec xs l h
              $\textbf{and}$  introsort xs l h <= sort_spec xs l h
  \end{lstlisting}
  \end{theorem}

@@

%   It contains the \emph{unguarded}
%   optimization: if parameter $G$ is false, we assume that we will hit a smaller element before underflowing the list, and thus can omit
%   the comparison $l<i$ in line~\ref{alg:insert:loophead}. In particular when element comparisons are cheap, i.e.~have cost similar to index comparisons,
%   omitting a comparison from the very inner loop can have a significant effect. Again, we show the version for pdqsort here, where the unguarded optimization
%   is used for non-leftmost partitions. The final insertion sort of introsort uses a similar optimization, where we switch to unguarded insertion once
%   the first threshold elements have been sorted.

  Again, we split the insertion sort algorithm into two smaller parts, which are proved separately via the following
  specification for \textsc{insert}:
  \begin{lstlisting}
    assert sorted xs[l..<i] \<and> l<=i \<and> i<|xs| \<and> (G \<or> xs[l-1] <= xs[i]);
    spec xs'. xs=$_{l,i+1}$xs' \<and> sorted xs[l..<i+1]
  \end{lstlisting}
  This captures the intuition that \textsc{insert} goes from a slice that is sorted up to index $i$ to one that is sorted up to index $i+1$.

@@

\section{Imperative Implementation}\label{sec:imp_impl}
  We have presented a refinement based approach to verify the introsort and pdqsort algorithms,
  including most optimizations we found in their libstdc++ and Boost implementations. However, the algorithms are still expressed
  as nondetermistic programs on functional lists and unbounded natural numbers.
  In this section, we use the Isabelle-LLVM framework~\cite{La19-llvm} to (semi-)automatically
  refine them to LLVM programs on arrays and 64 bit integers.

  \subsection{The Sepref Tool}
  The Sepref tool~\cite{La15,La19-llvm} symbolically executes an abstract program in the \is{nres}-monad,
  keeping track of refinements for every abstract variable to a concrete representation, which may use pointers to dynamically allocated memory.
  During the symbolic execution, the tool synthesizes an imperative Isabelle-LLVM program, together with a refinement proof.
  The synthesis is automatic, but usually requires some program-specific setup and boilerplate. For a detailed discussion of Sepref and Isabelle-LLVM,
  we refer the reader to~\cite{La15,La19-llvm}.

@@

  We now describe an approach towards solving this problem for Sepref.
  Abstractly, we model an array by the type \is{\<alpha> option list}\footnote{Here, \is$\\<alpha> option = None | Some \\<alpha>$ is Isabelle's option datatype,
  and \is{the (Some x) == x} is the corresponding selector function.},
  where \is{None} means that the array does currently not own the respective element.
  The abstract indexing operation then moves the
  element from the list to the result:
  \begin{lstlisting}
    move xs i == assert i<|xs| \and xs[i]~=None; return (the (xs[i]), xs[i:=None])
  \end{lstlisting}
%   where \is{the (Some x) = x} is the selector function of the option data type.
  As no memory is shared between the result and the array, we can show the following refinement:
  \begin{lstlisting}
    (%a i. (a[i],a), %xs i. move xs i) : oarr$^d$ \x nat$_{64}^k$ -> A \x oarr
  \end{lstlisting}
  where \is{oarr} is the relation between an array and an \is{\<alpha> option list}, and \is$A$ is the relation for the array elements.
  Note that this operation does not change the concrete array \is{a}. The movement of ownership is a purely abstract concept, which results
  in no implementation overhead.

  The transition from \is{\<alpha> list} to \is{\<alpha> option list} can typically be done in an additional refinement step,
  and thus does not obfuscate the actual correctness proofs, which are still done on plain \is{\<alpha> list}.
  Moreover, the \is{\<alpha> option list} representation is only required for subroutines where extracted array elements are actually visible.
  For example, we define an operation to compare two array elements:
  \begin{lstlisting}
    cmp_idxs xs i j == assert i<|xs| \<and> j<|xs|; return xs[i] < xs[j]
  \end{lstlisting}
  Inside this operation, we have to temporarily extract the elements $i$ and $j$ from the array,
  requiring an intermediate refinement step to \is{\<alpha> option list}.
  However, at the start and end of this operation, the array owns all its elements. For the whole operation, we thus get
  a refinement on plain arrays:
  \begin{lstlisting}
    (cmp_impl, cmp_idxs) : arr$^k$ \x nat$_{64}^k$ \x nat$_{64}^k$ -> bool
  \end{lstlisting}
  In our case, we only have to explicitly refine \textsc{insert} and \textsc{sift\_down}.
  The other subroutines use \is{cmp_idxs} and \is{swap} operations on plain lists.

@@

\section{Benchmarks}
%   We have verified the libstdc++ implementation of introsort, and the Boost implementation of pdqsort.
%   We benchmarked our verified implementations against the unverified originals.
%   As we did not verify the C++ code directly, re-implemented the algorithm using Isabelle-LLVM, we have to
%   test how close we came to the performance of the original.

@@

  \begin{figure}
  \plotpdqsort
  \plotintrosort
  \caption{Benchmarking our verified implementations against the unverified originals.
    For each element type, machine, and distribution, the value $(t_1/t_2-1)*s$ is shown,
    where $t_1$ is the slower time, $t_2$ is the faster time, and $s = 100$ if our implementation is slower,
    and $s=-100$ if the original implementation is slower. That is, a positive value $p$ indicates that our implementation is slower,
    requiring $100+p$ percent of the run time of the original. Analogously,
    a negative value $-p$ means that the original implementation is slower, requiring $100+p$ percent of our implementation's run time.
  }\label{fig:benchres}
  \end{figure}
